---
title: "Preprocessing Romeo Malette Plot Data"
author: "Harry Seely"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---

# Setup

```{r setup, include=FALSE}

start_time <- Sys.time()

#All chunks echo
knitr::opts_chunk$set(echo = TRUE)

```



# Get packages

```{r}

library(tidyverse)
library(here)
library(sf)
library(ggpubr)

#Import custom functions
source(here("A_preprocessing/preprocessing_utils.R"))

```


# Load allometric equations

Allometric equations from Lambert et al. (2003) and Ung et al. (2008)

Lambert et al: https://doi.org/10.1139/x05-112

Ung et al: https://doi.org/10.1139/X07-224

Allometric equation parameters can be downloaded at:
https://apps-scf-cfs.rncan.gc.ca/calc/en/biomass-calculator

```{r}

#Load allometric equation equation parameters and filter to Lambert et al. (2003)

eqs_dbh <- read_csv(here("data/ung_lambert_allometric_eqs_DBH_only.csv"))

eqs_dbh_ht <- read_csv(here("data/ung_lambert_allometric_eqs_DBH_Height.csv"))

```

Set main directories

```{r}

#Directory where clipped las files are stored
clipped_las_dir <- here("data/plot_point_clouds")

```

Load plot data

```{r}

#Load data

trees <- read_csv(here("data/romeo_malette/romeo_individual_tree_measurements.csv"))

sp_codes <- read_csv(here("data/romeo_malette/romeo_species_codes.csv"))

plots <- st_read(here("data/romeo_malette/plot_locations/MasterPlotLocationUTM_December6.shp"))

plot_info <- read_csv(here("data/romeo_malette/tblPlot.csv"))

#Join the species info with the trees df
trees <- left_join(trees, sp_codes, by = "SpeciesCode")
rm(sp_codes)

#Join plot info with plots and trees
trees <- left_join(trees, plot_info, by = "PlotKey")

#Rename PlotName as PlotID
trees <- trees %>% rename(PlotID = PlotName)

#ASSIGN A UNIQUE TREE ID across all trees
trees <- tibble::rowid_to_column(trees, "tree_ID")
```

# Examine tree data

Check for NAs in plot ID

```{r}

#check for NAs in plot Key or plot ID
table(is.na(trees$PlotKey))
table(is.na(trees$PlotID))

```

Check the sampling dates
```{r}

trees <- trees %>%
  mutate(samp_date = as.Date(MsrDate, format = "%Y-%m-%d")) %>%
  mutate(samp_year = year(samp_date))

sort(unique(trees$samp_date))

#Drop plots that were sampled in 2017 (2 plots)
trees <- trees %>%
  filter(samp_year != 2017)

sort(unique(trees$samp_year))

```



# Clean tree data

Check status codes

```{r}
#What percent of measured trees are dead?
round(table(trees$StatusCode)/nrow(trees)*100,1)

```

Clean trees df and remove dead trees

```{r}
#Remove dead trees (only estimating live biomass for now)
trees <- trees %>% filter(StatusCode == "L"| StatusCode == "V")

#Rename vars
trees <- trees %>% rename(species = Common,
                          height = TotalHeight,
                          dbh = DBH)


```

# Get species in plots

```{r}

trees %>%
  group_by(species) %>%
  summarize(
            n = n(),
            perc = round(n()/nrow(trees)*100, 2)
  ) %>%
  arrange(desc(n))


```




Rename tree species to match allometric equations

```{r}

#Paper Birch as White Birch 
trees$species[trees$species == "Paper Birch"] <- "White Birch"

#American Larch as Tamarack Larch
trees$species[trees$species == "American Larch"] <- "Tamarack Larch"

#Bitternut Hickory as Hickory
trees$species[trees$species == "Bitternut Hickory"] <- "Hickory"

#Pitch Pine as Coniferous (since there is only 1 across entire dataset, 
#...not worth gathering exact allometric equation parameters for)
trees$species[trees$species == "Pitch Pine"] <- "Conifers"

#Eastern Red Cedar as Coniferous (since there is only 1 across entire dataset, 
#...not worth gathering exact allometric equation parameters for)
trees$species[trees$species == "Eastern Red Cedar"] <- "Conifers"

#Assign trees without ID species var to "All" to use generic biomass eq
trees$species[is.na(trees$species)] <- "All"

#Capitalize allometric eq species names for both lists of parameters
eqs_dbh$Species_en <- str_to_title(eqs_dbh$Species_en)
eqs_dbh_ht$Species_en <- str_to_title(eqs_dbh_ht$Species_en)

#Check that all species names in data coreespond to those in allometric eqs
print("All species in data are in allometric eqs:")
table(unique(trees$species) %in% unique(eqs_dbh$Species_en))
table(unique(trees$species) %in% unique(eqs_dbh_ht$Species_en))

```

Check DBH values for NAs
```{r}

#Check for NA values in height and DBH cols
print("Trees with/without DBH measurements")
print(table(!is.na(trees$dbh)))

#Remove trees that have no DBH measurement
trees <- trees %>% filter(!is.na(dbh))

#Check for NA values in height
print("Trees with/without height measurements")
print(table(!is.na(trees$height))/nrow(trees))

#Check for NAs in species
table(is.na(trees$species))

```

Inspect height values 

```{r}

trees %>%
  ggplot(aes(x = height)) +
  geom_histogram() +
  ggtitle("Histogram of Tree Heights") +
  theme_minimal()

```

Pivot allometric equations to wide format

```{r}
eqs_dbh_wide <- eqs_dbh %>% select(Species_en, Component_en, a, b) %>%
  pivot_wider(names_from = Component_en, values_from = c(a, b))

eqs_dbh_ht_wide <- eqs_dbh_ht %>% select(Species_en, Component_en, a, b, c) %>%
  pivot_wider(names_from = Component_en, values_from = c(a, b, c))

```

# Calculate tree component biomass

*NOTE: biomass is calculated in kilograms (kg)

Compare biomass values calculate with and without height measurements

```{r}

# Extract trees that have height measurements
trees_yes_ht <- trees %>% 
                  filter(!is.na(height))

trees_biomass_yes_ht <- trees_yes_ht %>% 
                            select(-contains(c("a_", "b_", "c_"))) %>%
                            left_join(eqs_dbh_ht_wide, by = c("species" = "Species_en")) %>%
                            get_tree_comp_biomass(use_height = TRUE) %>%
                            mutate(allometric_input = "height & dbh")

trees_biomass_no_ht <- trees_yes_ht %>% 
                            select(-contains(c("a_", "b_", "c_"))) %>%
                            left_join(eqs_dbh_wide, by = c("species" = "Species_en")) %>%
                            get_tree_comp_biomass(use_height = FALSE) %>%
                            mutate(allometric_input = "dbh only")

allo_comparison <- bind_rows(trees_biomass_yes_ht, trees_biomass_no_ht) 

allo_comparison %>% 
            ggplot(aes(y = wood_biomass, fill = allometric_input)) +
            geom_boxplot() +
            theme_minimal()  

agb_allo_comparison <- allo_comparison %>%
  select(tree_ID, tree_AGB, allometric_input) %>%
  pivot_wider(names_from = allometric_input, values_from = tree_AGB) %>%
  mutate(agb_diff = `height & dbh` - `dbh only`)

agb_allo_comparison %>%
  ggplot(aes(x = `dbh only`, y = `height & dbh`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  stat_cor() +
  theme_minimal() +
  ggtitle("AGB Allometric Comparison with and without Height Measurements") +
  xlab("AGB (dbh only)") +
  ylab("AGB (height & dbh)")


agb_allo_comparison %>%
  summarize(
    mean_diff = mean(agb_diff),
    sd_diff = sd(agb_diff),
    min_diff = min(agb_diff),
    max_diff = max(agb_diff)
  ) %>%
  print()

rm(trees_biomass_yes_ht, 
  trees_biomass_no_ht, 
  allo_comparison, 
  agb_allo_comparison, 
  trees_yes_ht)

```


Calculate biomass using DBH-only allometries

```{r, echo=TRUE}

trees <- trees %>% left_join(eqs_dbh_wide, by = c("species" = "Species_en"))

trees <- get_tree_comp_biomass(trees, use_height = FALSE)

head(trees)

```

Check for NAs

```{r, echo=FALSE}

print("Foliage"); table(!is.na(trees$foliage_biomass))
print("Bark"); table(!is.na(trees$bark_biomass))
print("Branches"); table(!is.na(trees$branches_biomass))
print("Wood"); table(!is.na(trees$wood_biomass))
print("Total AGB"); table(!is.na(trees$tree_AGB))

```

Check range of biomass values
```{r, echo=TRUE}

#Check vals
print("Foliage"); range(trees$foliage_biomass)
print("Bark"); range(trees$bark_biomass)
print("Branches"); range(trees$branches_biomass)
print("Wood"); range(trees$wood_biomass)
print("Total AGB"); range(trees$tree_AGB)


```

# Aggregate biomass calculations for each plot

Convert biomass to biomass to metric tonnes per ha (1t == 1000 kg, per hectare)

Each Romeo Malette plot has a radius of 11.28m, so an area of 399.7m^2

1m^2 == 0.0001 ha

So each plot is 0.03997 ha

```{r}

#How many plots have tree data?
length(unique(trees$PlotID))

#Check for NA values in vars
table(is.na(trees$PlotID))

#Get plot IDs
unique(trees$PlotID)

#Summarize biomass by component for each plot
biomass_by_plot <- trees %>% group_by(PlotID) %>% summarise(
  bark_total = sum(bark_biomass),
  branch_total = sum(branches_biomass),
  foliage_total = sum(foliage_biomass),
  wood_total = sum(wood_biomass),
  total_AGB = sum(tree_AGB))

plot_area_in_ha <- 0.03997

#Divide by 1000 for tonnes conversion, then by plot area in ha (0.03997) to get tonnes/ha
biomass_by_plot <- biomass_by_plot %>% mutate(total_Mg_ha = total_AGB/1000/0.03997)
biomass_by_plot <- biomass_by_plot %>% mutate(branch_Mg_ha = branch_total/1000/0.03997)
biomass_by_plot <- biomass_by_plot %>% mutate(foliage_Mg_ha = foliage_total/1000/0.03997)
biomass_by_plot <- biomass_by_plot %>% mutate(bark_Mg_ha = bark_total/1000/0.03997)
biomass_by_plot <- biomass_by_plot %>% mutate(wood_Mg_ha = wood_total/1000/0.03997)

#Check for NAs
print("Checking NAs:")
check_df_for_nas(biomass_by_plot)


```

# Summarize biomass in RMF and other plot stats

```{r, echo=FALSE}

#Get the range and mean of AGB
print(range(biomass_by_plot$total_Mg_ha))
print(mean(biomass_by_plot$total_Mg_ha))

total_agb_rmf <- sum(biomass_by_plot$total_AGB)

#Summarize component proportions
biomass_comp_prop_summary <- biomass_by_plot %>%
      select(c("foliage_total",
               "bark_total",
               "branch_total",
               "wood_total")) %>%
      pivot_longer(cols = everything(),
                   names_to = "comp",
                   values_to = "biomass") %>%
      group_by(comp) %>%
      summarize(agb_prop = round(sum(biomass) / total_agb_rmf, 2) * 100)
               
print(biomass_comp_prop_summary)



```


# Prep Data For Deep Learning

```{r, echo=TRUE}

#Create new df for data ready for dl
#Include spectral values in df
dl_df <- biomass_by_plot %>%
    select(PlotID, contains("_Mg_ha"))

#Check that plotids match with las filenames
fnames <- list.files(clipped_las_dir, full.names=FALSE)
fnames <- str_replace(fnames, ".las", "")
fnames <- str_replace(fnames, "plot_", "")

print("Checking that plot IDs match with las filenames:")
table(dl_df$PlotID %in% fnames)

#NA check and removal
na_df <- check_df_for_nas(dl_df, return_col_names = T)
na_cols <- names(na_df)
print(paste("The following columns have NAs:", na_cols))

#Check that each biomass component values makes sense across all datasets

#HISTOGRAM
dl_df %>%
    select(contains("Mg_ha")) %>%
    pivot_longer(cols = contains("Mg_ha"), names_to = "component", values_to = "biomass") %>%
        mutate(component = str_remove(component, "_Mg_ha")) %>%
        mutate(component = factor(component, levels = c("foliage", "bark", "branch", "wood", "total"))) %>%
    ggplot(aes(x = biomass, fill = component)) +
    geom_histogram() +
    facet_wrap(~component, scales="free") +
    ggtitle("Histogram of Biomass Components per Hectare") +
    theme_minimal()

#BOXPLOT
dl_df %>%
    select(contains("Mg_ha")) %>%
    pivot_longer(cols = contains("Mg_ha"), names_to = "component", values_to = "biomass") %>%
        mutate(component = str_remove(component, "_Mg_ha")) %>%
        mutate(component = factor(component, levels = c("foliage", "bark", "branch", "wood", "total"))) %>%
    ggplot(aes(x = biomass, fill = component)) +
      geom_boxplot() +
      coord_flip() +
      facet_wrap(~component, scales="free") +
      ggtitle("Boxplot of Biomass Components per Hectare") +
      theme_minimal()

#Check the contribution to tree biomass from each component -----
foliage_perc <- round(sum(dl_df$foliage_Mg_ha)/sum(dl_df$total_Mg_ha)*100, 2)
bark_perc <- round(sum(dl_df$bark_Mg_ha)/sum(dl_df$total_Mg_ha)*100, 2)
branch_perc <- round(sum(dl_df$branch_Mg_ha)/sum(dl_df$total_Mg_ha)*100, 2)
wood_perc <- round(sum(dl_df$wood_Mg_ha)/sum(dl_df$total_Mg_ha)*100, 2)

paste("Total Foliage % of tree biomass:", foliage_perc)
paste("Total Bark % of tree biomass:", bark_perc)
paste("Total Branches % of tree biomass:", branch_perc)
paste("Total Wood % of tree biomass:", wood_perc)

foliage_perc + branch_perc + wood_perc + bark_perc


```

Convert biomass comps to z-score

```{r}

#Get z score for each component biomass (Mg/ha)
dl_df$foliage_z <- convert_to_z_score(dl_df$foliage_Mg_ha)
dl_df$bark_z <- convert_to_z_score(dl_df$bark_Mg_ha)
dl_df$branch_z <- convert_to_z_score(dl_df$branch_Mg_ha)
dl_df$wood_z <- convert_to_z_score(dl_df$wood_Mg_ha)
dl_df$total_z <- convert_to_z_score(dl_df$total_Mg_ha)

```


# Split data into separate test and train/val sets, then generate k-fold train-val splits using statified sampling

Set params
```{r}
#Set proportion for test, and then relative proportions for train and val
test_prop <- 0.17
val_prop <- 0.13 
train_prop <- 1 - test_prop - val_prop

#Set relative train - val props
rel_train_prop <- train_prop/(1 - (test_prop))

#Get total number of points for train and val/test sets based on propertion of total data
n_train <- round(nrow(plots) * train_prop, 0)
n_val <- round(nrow(plots) * (1 - train_prop)/2, 0)
n_test <- nrow(plots) - n_train - n_val

stopifnot(nrow(plots) == n_train + n_val + n_test)

cat(train_prop, val_prop, test_prop)
cat(n_train, n_val, n_test)

#Set stratified sampling params
strat_var <- "total_Mg_ha"
n_bins <- 7

```

Split dataset into separate train and test sets using stratified sampling

```{r}

#Assign plots to bins based on total_Mg_ha
dl_df['bins'] <-  ntile(dl_df[strat_var], n=n_bins)

#Get the train/val set
set.seed(65)
test_df <- dl_df %>%
  group_by(bins) %>%
  slice_sample(prop = test_prop) %>%
  mutate(split = "test")

#Get the train/val df
train_val_df <- dl_df %>%
  anti_join(test_df, by = "PlotID")


hist(test_df$total_Mg_ha, main = "Test set biomass distribution", xlab = "Biomass (Mg/ha)")

```

Define function to divide dataset into k-folds

```{r}

split_dataset <- function(i, df, rel_train_prop, id_col){

  #Get the train set
  set.seed(i)
  train_df <- df %>%
    group_by(bins) %>%
    slice_sample(prop = rel_train_prop) %>%
    mutate(split = "train")

  #Get val set
  val_df <- df %>%
      anti_join(train_df, by = "PlotID") %>%
      mutate(split = "val")

  #Combine
  fold_df <- train_df %>%
    bind_rows(val_df) %>%
    ungroup()

  #Rename the column for split
  names(fold_df) <- str_replace(names(fold_df), "split", paste0("strat_fold_", i))

  #Return the PlotID with the fold col
  out_df <- fold_df %>%
    select(all_of(id_col), starts_with("strat_fold_"))

  return(out_df)
}

#Set parameters for train/test/val splits
n_folds <- 5

#Split datasets
out_strat_folds <- lapply(X=1:n_folds,
            FUN=split_dataset,
            df=train_val_df,
            rel_train_prop=rel_train_prop,
            id_col="PlotID")

#Join all output dfs in list using PlotID col
strat_fold_dfs <- Reduce(function(x, y) merge(x, y, by = "PlotID"), out_strat_folds)

#Join splits df with dl df
dl_df <- dl_df %>%
  left_join(strat_fold_dfs, by = "PlotID")

#Populate NA rows in strat fold cols with test set
for (f in 1:n_folds){
  fold_nm <- paste0("strat_fold_", f)
  dl_df[is.na(dl_df[fold_nm]),][fold_nm] <- "test"
}

cv_split_summary <- dl_df %>% 
  select(contains("strat_fold_")) %>%
  pivot_longer(cols = contains("strat_fold_"), names_to = "strat_fold", values_to = "set") %>%
  group_by(strat_fold, set) %>%
  summarize(n = n()) %>%
  mutate(prop = round(n/nrow(dl_df) * 100, 1)) 

print(head(cv_split_summary))
print("Check that splits sum to 100%")
print(sum(round(cv_split_summary$prop/n_folds, 0)))

#Check for duplicate PlotIDs
print("Checking for duplicate PlotIDs")
stopifnot(sum(duplicated(dl_df$PlotID)) == 0)

```


```{r}

#Visualize distributions of test vs. train-val sets
dl_df %>%
  select(PlotID, starts_with("strat_fold")) %>%
  slice_head(n = 20)

```

```{r}

#Visualize distributions of test vs. train-val sets
dl_df %>%
  select(total_Mg_ha, starts_with("strat_fold")) %>%
  pivot_longer(cols = starts_with("strat_fold"), names_to = "strat_fold", values_to = "set") %>%
  ggplot(aes(x = total_Mg_ha, fill = set)) +
  facet_wrap(~strat_fold) +
  geom_histogram(position = "dodge", bins = 20) +
  ggtitle("Distribution of test and train-val sets") +
  theme_minimal()

```

# Relocate las files to new folder with split info

```{r}

#Get names of las files in clip dir
las_fnames <- list.files(clipped_las_dir)

#Ensure las files correspond with plot ids
fname_filt <- str_replace(las_fnames, ".las", "") %in% paste0("plot_", dl_df$PlotID)
las_fnames <- las_fnames[fname_filt]

print("Checking that las files correspond with plot IDs:")
table(str_replace(las_fnames, ".las", "") %in% paste0("plot_", dl_df$PlotID))

```

# Export results

```{r, echo=TRUE}

#Final NA check
check_df_for_nas(dl_df)

#Export plot data prepped for DL (i.e., Z scored)
write_csv(dl_df, here("data/biomass_labels.csv"))

#Export the plot locations with updated spatial cross validation folds
st_write(plots, here("data/plot_locations.gpkg"), append = FALSE)

```


# Run time 

```{r, echo=TRUE}

end_time <- Sys.time()

# Calculate elapsed time in seconds
elapsed_time <- as.numeric(difftime(end_time, start_time, units = "secs"))

# Convert elapsed time to hours, minutes, and seconds
elapsed_hours <- floor(elapsed_time / 3600)
elapsed_minutes <- floor((elapsed_time %% 3600) / 60)
elapsed_seconds <- elapsed_time %% 60

# Print the elapsed time in hours, minutes, and seconds
cat("Elapsed time:", elapsed_hours, "hours,", elapsed_minutes, "minutes,", elapsed_seconds, "seconds.")
  
```

# Reproducibility

```{r, echo=TRUE}

Sys.info()

sessionInfo()

```
