---
title: "R Notebook"
output: html_notebook
---
# Get packages

```{r}

library(lidR)
library(sf)
library(here)
library(future)
library(parallel)
library(future.apply)
library(dplyr)

```

#Set global vars

```{r}
PLOT_RADIUS <- 11.28

SAMPLE_PTS_FPATH <- here("data/unlabeled_samples.gpkg")

TILES_DIR <- "E:/RMF/RMF_SPL100/LAS_Classified_Point_Clouds_Normalized"

CLIP_LAS_OUT_DIR <- "E:/RMF/RMF_SPL100/clipped_unlabeled_samples"

NCORES <- 10

```

Read the plot locations

```{r}

#Read the unlabeled sample points
sample_pts <- st_read(SAMPLE_PTS_FPATH)
```

Read las catalog

```{r}
ctg <- lidR::readLAScatalog(TILES_DIR)
ctg
```
Modify some LASCatalog options to speed up clipping 

```{r}

#Only read xyz coords, no other lidar attributes needed
opt_select(ctg) <- "xyz"

#Do not plot progress
opt_progress(ctg) <- FALSE

#Do not stop if running into an erro
opt_stop_early(ctg) <- FALSE

```


Clear existing files from output directory (if specified)

```{r}

#Delete existing files in directory (input required)
rm_files <- readline(prompt="Remove existing files (TRUE/FALSE): ")

if(rm_files == "TRUE"){
  unlink(CLIP_LAS_OUT_DIR, recursive = TRUE)
}


#Read the existing clipped files and remove them from the sample_pts df
clipped_fnames <- list.files(CLIP_LAS_OUT_DIR, full.names = F)
clipped_ids <- as.integer(gsub(".las", "", clipped_fnames))

#Remove the clipped files from the sample_pts df
print(paste("Removing", 
            length(clipped_ids), 
            "clipped files from the sample_pts df"))
            
sample_pts <- sample_pts %>% 
  filter(!ID %in% clipped_ids)



```

Get a vector of xy coords from df

```{r}

#Extract x and y coords 
sample_pts <- sample_pts %>% 
  mutate(x = st_coordinates(geom)[,1],
         y = st_coordinates(geom)[,2]) %>%
  select(x, y, ID) %>%
  st_drop_geometry() %>% 
  as.matrix() %>% 
  unname()

```


Define function to clip points from ctg

```{r}

clip_ctg <- function(x_y_id, ctg, rad, out_dir){
  
  #x_y_id: matrix with the values x, y, and ID in order
  #ctg: LASCatalog object
  #rad: radius of the circle from which to clip the ctg
  #out_dir: output directory for the clipped file
  
  tryCatch({
    
  #Set output fpath for ctg
  opt_output_files(ctg) <- file.path(out_dir, x_y_id[3])
  
  #Test clip circle for one plot
  clip_circle <- lidR::clip_circle(ctg, 
                                   xcenter=x_y_id[1], 
                                   ycenter=x_y_id[2], 
                                   radius=rad)
  
  }, error = function(e) {
    
      #Report error
      print(paste("Error for sample ID:", x_y_id[3]))
      print(e)
  })
}

```

Test fn on plot

```{r}

#Grab a test row
test_row <- sample_pts[1,]

#Test function using a single plot
clip_ctg(x_y_id=test_row, 
            ctg=ctg, 
            rad=PLOT_RADIUS, 
            out_dir=CLIP_LAS_OUT_DIR)

rm(test_row)

```

Test function for 10 rows using apply

```{r}

#Grab a test row
test_matrix <- sample_pts[1:10,]

#Test function using several plots
apply(test_matrix,
      MARGIN=1, 
      FUN=clip_ctg, 
      #Func Args
      ctg=ctg, 
      rad=PLOT_RADIUS, 
      out_dir=CLIP_LAS_OUT_DIR)

rm(test_matrix)

```

Function to clip rasters in parallel

```{r}

clip_ctg_par <- function(x_y_id_matrix, ctg, rad, out_dir, ncores){
  #x_y_id_matrix: matrix with x, y, and ID as cols and each point as rows
  
  #Initiate parallel processing
  cl <- parallel::makeCluster(NCORES, timeout = 60 * 5)
  plan(cluster, workers = cl)
  
  #Clip the rasters
  future_apply(x_y_id_matrix, 
               MARGIN=1, 
               FUN=clip_ctg, 
               #Func Args
               ctg=ctg, 
               rad=rad, 
               out_dir=out_dir,
               future.seed=TRUE)
  
  #Stop the cluster
  parallel::stopCluster(cl)
  
}

```


Clip laz files in parallel

```{r}

#Get start time
start_time <- Sys.time()

clip_ctg_par(x_y_id_matrix=sample_pts, 
             ctg=ctg, 
             rad=PLOT_RADIUS, 
             out_dir=CLIP_LAS_OUT_DIR, 
             ncores=NCORES)

#Report runtime
end_time <- Sys.time()

#Sumamrize runtime
elapsed_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
elapsed_hours <- floor(elapsed_time / 3600)
elapsed_minutes <- floor((elapsed_time %% 3600) / 60)
elapsed_seconds <- elapsed_time %% 60

#Calculate time per plot
s_per_plot <- elapsed_time / nrow(sample_pts)

#Extrapolate to one million samples
s_per_million <- s_per_plot * 1e6

#Print as hours
h_per_million <- s_per_million / 60 / 60

#Print to console
print(paste("For", nrow(sample_pts), "files using", NCORES, "cores:"))

print(paste("Elapsed time:", elapsed_hours, "hours,", 
            elapsed_minutes, "minutes,", elapsed_seconds, "seconds for",
            nrow(sample_pts), "plots."))

print(paste("Estimated time of", round(h_per_million, 2), "hours for 1M plots"))

```

